detect patches

### 2024-01-30 scripts/classify_patches.Rmd
writing fn to calc landscape metrics

```{r}
library(terra)
```

@@

patcheR <- function(x){

  # read in file
  r <- rast(x)
  taxon <- gsub('1k.*$', '', basename(x))
  pout <- '../results/patches'
  fsize <- file.info(x)$size/1e9

  # mask raster values < 0.8
  r <- terra::mask(r, ifel(r < 0.80, NA, r))

    # burn away rivers, and burn patches to hydrologic units ~ populations
  r <- terra::mask(r, terra::crop(hydr_bound, r), inverse = TRUE)
  r <- terra::mask(r, terra::crop(hydr_ftr, r), inverse = TRUE)

  # detect patches - this scales to the extent of analysis. #
  # Areas need to be aggregateed for computatations to be run in meaningful amount of
  # time

  if(fsize < 0.3){r1 <- terra::aggregate(r, 2, fun = 'mean')} else
    if(fsize < 0.5){r1 <- terra::aggregate(r, 3, fun = 'mean')} else 
      if(fsize < 0.7){r1 <- terra::aggregate(r, 4, fun = 'mean')} else
        if(fsize > 0.9){r1 <- terra::aggregate(r, 5, fun = 'mean')}

  pat <- terra::patches(r1,  directions = 4, allowGaps = FALSE)

  # remove patches < 5 acres
  sizes <- terra::zonal(cellSize(pat, unit="ha"), pat, sum, as.raster=TRUE)
  size_mask <- terra::ifel(sizes < 4.0470, NA, sizes)
  r1 <- mask(r1, size_mask)

  # resample the patches to the original raster resolution
  terra::resample(r1, r, threads = 16, method = 'near',
           filename = file.path(pout, paste0(taxon, '.tif')))

  terra::tmpFiles(current = FALSE, orphan = TRUE, old = TRUE, remove = TRUE)

@@

lapply(f[4:355], patcheR) # start 9:05

@@

# euclidean nearest neighbor distance (enn) - (Patch)
## (modification) log(enn), sqrt(enn)
### Expectation: the number of colonized predicted patches decrease with increasing distance.
#### why: long-distance dispersal is limited from colonized to suitable patches. 
##### test: glm(colonization(0|1) ~ enn,  family = binomial)

# nearest neighbor distance (enn_mn) - (Class) # note other method of aggregation can be used too!
## (modification) log(enn_mn), sqrt(enn_mn)
### Expectation: the number of colonized predicted patches increases with increasing distance.
#### why: this species is adept at long distance dispersal
##### test: glm(colonization(0|1) ~ enn_mn,  family = binomial)

@@

# Core Area Index (cai)  
## (modification) arithmetic mean of core area cells
### Expectation: higher cai higher abundance
#### why: better within patch distribution, and a persistent 'source' to 'sink' peripheral cells. 
##### test: glmm?(% cover ~ cai, correlation = nlme::corExp(Long ~ Lat, nugget=T), family = quasibinomial)

# perimeter-area ratio (para)
## (modification) none
### Expectation: lower para higher abundance
#### why: areas with 'large edges' will suffer dispersal limitations throughout them, as many seeds will go to unsuitable habitat.
##### test: glmm?(% cover ~ para, correlation = nlme::corExp(Long ~ Lat, nugget=T), family = quasibinomial)

# fractal dimension index (frac)
## (modification) none
### Expectation: lower fractal dimensions (less complex patch shapes) have higher abundances. 
#### why: components of patches with complex shapes may suffer dispersal limitation. 
##### test: glmm?(% cover ~ frac, correlation = nlme::corExp(Long ~ Lat, nugget=T), family = quasibinomial)

>>>>

```{r}
library(terra)
library(landscapemetrics)
```

Create coarse patches for 'meta-populations'

@@

lapply(f, patcheR) # roughly 72 hours.

abvi <- rast('../results/patches/Abronia_villosa.tif')

gd <- '/media/steppe/hdd/Geospatial_data'
hydr_bound <- rast( file.path(gd, 'WBD_rast', 'hu10R-line.tif') )

landscapR <- function(x, pout, pRout){

  x <- terra::ifel(x > 0.75, 1, NA)
  x <- terra::mask(x, f_wbd, inverse = TRUE)
  landscape <- landscapemetrics::get_patches(x, directions = 4)
  calcs <- landscapemetrics::calculate_lsm(x, what = c('lsm_c_enn_mn', 'lsm_c_enn_cv',
                "lsm_p_enn", "lsm_p_enn", "lsm_p_cai", "lsm_p_para", "lsm_p_frac"), 
              neighbourhood = 4, directions = 4)

  p <- '../results/'
  write.csv(
    calcs[calcs$level == 'patch', c('id', 'metric', 'value')],
    filename = paste0(p, pout, '/', taxon, '-', 'patch_vals.csv'),  row.names = F
  )

  write.csv(
    calcs[calcs$level == 'class', c('id', 'metric', 'value')],
    filename = paste0(p, pout, '/', taxon, '-', 'class_vals.csv'), row.names = F
  )

  writeRaster(landscape[["layer_1"]][["class_1"]], 
    filename = paste0(p, pRout, '/', taxon, '-', 'patchIDS.tif')
  )

@@

landscapR(f, pout = 'patch_metrics', pRout = 'fine_patches')

@@

# euclidean nearest neighbor distance (enn) - (Patch) {lsm_p_enn}
## (modification) log(enn), sqrt(enn)
### Expectation: the number of colonized predicted patches decrease with increasing distance.
#### why: long-distance dispersal is limited from colonized to suitable patches. 
##### test: glm(colonization(0|1) ~ enn,  family = binomial)

# nearest neighbor distance (enn_mn) - (Class) {lsm_c_enn_mn} # note other method of aggregation can be used too!
## (modification) log(enn_mn), sqrt(enn_mn)
### Expectation: the number of colonized predicted patches increases with increasing distance.
#### why: this species is adept at long distance dispersal
##### test: glm(colonization(0|1) ~ enn_mn,  family = binomial)

@@

# Core Area Index (cai)  {lsm_p_cai}
## (modification) arithmetic mean of core area cells
### Expectation: higher cai higher abundance
#### why: better within patch distribution, and a persistent 'source' to 'sink' peripheral cells. 
##### test: glmm?(% cover ~ cai, correlation = nlme::corExp(Long ~ Lat, nugget=T), family = quasibinomial)

# perimeter-area ratio (para) {lsm_p_para}
## (modification) none
### Expectation: lower para higher abundance
#### why: areas with 'large edges' will suffer dispersal limitations throughout them, as many seeds will go to unsuitable habitat.
##### test: glmm?(% cover ~ para, correlation = nlme::corExp(Long ~ Lat, nugget=T), family = quasibinomial)

# fractal dimension index (frac) {lsm_p_frac}
## (modification) none
### Expectation: lower fractal dimensions (less complex patch shapes) have higher abundances. 
#### why: components of patches with complex shapes may suffer dispersal limitation. 
##### test: glmm?(% cover ~ frac, correlation = nlme::corExp(Long ~ Lat, nugget=T), family = quasibinomial)
